Norwegian citizen science data exploration

### 2018-04-16 paper.Rmd
Generate pseudoabsences

  ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
  spring <- as.Date("3-15",  format = "%m-%d") ## From Mar 15 
  summer <- as.Date("5-22",  format = "%m-%d")  ## From May 22
  fall <- as.Date("9-11",  format = "%m-%d")  ## From Sep 11
  winter <- as.Date("11-18",  format = "%m-%d")  ## From Nov 18

@@

A test data exploration of some [Norwegian Species Observation Service](https://www.artsobservasjoner.no) data, obtained through [GBIF](https://www.gbif.org) (N = `r nrow(ao_data)`).

Filtering on `r taxon['canonicalName']` observations, plotted below.

```{r map_per_species, echo=FALSE, message=FALSE}
  ggplot(as.data.frame(ao_data)) +
    annotation_map(as.data.frame(country_map), fill="gray70", colour="gray70") +
    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=.5, shape=20) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    facet_wrap(~ species)

@@

Map for each `r taxon['canonicalName']` species in Norway

@@

  gplot(ao_data_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") + 
    labs(title="Total observation density") +
    labs(subtitle=taxon['canonicalName']) +
    labs(caption = paste("Density map for all", taxon['canonicalName'], "observations in Norway, rasterized to", gridsizeinkm, "x", gridsizeinkm, "km cells."))

@@

## Human population density
```{r map_human_population, messages=FALSE, echo=FALSE}
  pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')

  gplot(pop_unrasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")
```

Log human population per square km in norway

@@

gplot(pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")

@@

The log human population per `r gridsizeinkm` x `r gridsizeinkm` km cs cell in Norway.

@@

The number of `r taxon['canonicalName']` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells. R² of the linaer regression is `r summary(lm(observations ~ population, data=num_vs_pop))$r.squared`.

@@

gplot(num_vs_pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations per capita")
```

The log number of `r taxon['canonicalName']` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.

@@

gplot(num_vs_pop_rasterized_LM) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations relative to regression with human population")
```

The log number of `r taxon['canonicalName']` observations in Norway divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation),  within `r gridsizeinkm` x `r gridsizeinkm` km cells.

>>>>

  ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
  spring <- as.Date("2012-03-15",  format = "%Y-%m-%d") ## From Mar 15 
  summer <- as.Date("2012-05-22",  format = "%Y-%m-%d")  ## From May 22
  fall <- as.Date("2012-09-11",  format = "%Y-%m-%d")  ## From Sep 11
  winter <- as.Date("2012-11-18",  format = "%Y-%m-%d")  ## From Nov 18

@@

A test data exploration of some [Norwegian Species Observation Service](https://www.artsobservasjoner.no) data: observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, obtained through [GBIF](https://www.gbif.org) (N = `r nrow(ao_data)`).

## Observations

```{r map_per_species, echo=FALSE, message=FALSE}
  ggplot(subset(as.data.frame(ao_data),specificEpithet!="")) +
    annotation_map(as.data.frame(country_map), fill="white", colour="white") +
    geom_point(aes(x=decimalLongitude, y=decimalLatitude), color="darkgreen", alpha=.05, size=.25, shape=20) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    facet_wrap(~ species) + 
    ggtitle("Observations per species")

@@

Map of observations of each species within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway.

```{r map_per_season, echo=FALSE, message=FALSE}
  ggplot(as.data.frame(ao_data)) +
    annotation_map(as.data.frame(country_map), fill="white", colour="white") +
    geom_point(aes(x=decimalLongitude, y=decimalLatitude), color="darkgreen", alpha=.05, size=.25, shape=20) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    facet_wrap(~ season) + 
    ggtitle("Observations per season")

Map of all observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway, per season.

@@

  gplot(ao_data_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
    scale_fill_distiller(palette = "Spectral", na.value=NA, direction = -1) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Observations (log)") + 
    ggtitle("Observation density")

@@

Density map for all observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.

```{r rastermap_observations_seasonal, message=FALSE, echo=FALSE}

  # ao_data_rasterized_winter <- raster::rasterize(spTransform(subset(ao_data, season=='Winter'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
  # ao_data_rasterized_spring <- raster::rasterize(spTransform(subset(ao_data, season=='Spring'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
  # ao_data_rasterized_summer <- raster::rasterize(spTransform(subset(ao_data, season=='Summer'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
  # ao_data_rasterized_fall <- raster::rasterize(spTransform(subset(ao_data, season=='Fall'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')

  # ao_season_stack <- stack(ao_data_rasterized_winter, ao_data_rasterized_spring, ao_data_rasterized_summer, ao_data_rasterized_fall)
  # ao_season_stack <- log(ao_season_stack)
  # names(ao_season_stack) <- c('Winter','Spring','Summer','Fall')
  # 
  # levelplot(ao_season_stack,
  #     col.regions=colorRampPalette(rev(brewer.pal(5,"Spectral"))),
  #     scales=list(draw=FALSE ))

```{r rastermap_observations_seasonal_exp, message=FALSE, echo=FALSE}
#   
# 
#   
#   
  # ao_data_rasterized_winter <- ratify(ao_data_rasterized_winter)
  # rat <- levels(ao_data_rasterized_winter)[[1]]
  # rat$season <- "Winter"
  # levels(ao_data_rasterized_winter) <- rat
  # 
  # ao_data_rasterized_spring <- ratify(ao_data_rasterized_spring)
  # rat <- levels(ao_data_rasterized_spring)[[1]]
  # rat$season <- "Spring"
  # levels(ao_data_rasterized_spring) <- rat
  # 
  # ao_data_rasterized_summer <- ratify(ao_data_rasterized_summer)
  # rat <- levels(ao_data_rasterized_summer)[[1]]
  # rat$season <- "Summer"
  # levels(ao_data_rasterized_summer) <- rat
  # 
  # ao_data_rasterized_fall <- ratify(ao_data_rasterized_fall)
  # rat <- levels(ao_data_rasterized_fall)[[1]]
  # rat$season <- "Fall"
  # levels(ao_data_rasterized_fall) <- rat
  # 
  # 
  # ao_season_stack <- stack(ao_data_rasterized_winter, ao_data_rasterized_spring, ao_data_rasterized_summer, ao_data_rasterized_fall)

#   
#   
#   
   # View(as.data.frame(ao_data_rasterized_winter))
#   
# 
  # gplot(ao_data_rasterized_winter) +
  #   geom_tile(aes(fill = log(value))) +
  #   facet_wrap(~ name) +
  #   scale_fill_gradient(low = 'white', high = 'blue') +
  #   coord_equal()
# 
# 
#   ao_data_rasterized_hm<- raster::rasterize(spTransform(ao_data, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season', na.rm=TRUE)
#     
#     
#     
#     
# 
  # country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
#   
# gplot(ao_season_stack) +
#     coord_fixed() +
#     annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
#     geom_tile(aes(fill=log(value)), alpha=1) +
#     annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
#     scale_fill_distiller(palette = "Spectral", na.value=NA) +
#     guides(color = guide_legend(override.aes = list(size=4))) +
#     facet_wrap(~ layer) +
#     # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
#     xlab("Longitude") +
#     ylab("Latitude")

## Human population

```{r map_human_population, messages=FALSE, echo=FALSE}
  pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')

  gplot(pop_unrasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA, direction=-1) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Inhabitants (log)") +
    ggtitle("Human population density")
```

Log human population per 10 x 10 km cell in norway

@@

gplot(pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA, direction=-1) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Inhabitants (log)") +
    ggtitle("Human population density")

@@

The log human population in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.

@@

The number of observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` per `r gridsizeinkm` x `r gridsizeinkm` km cells, versus the human population in those cells. R² of the linear regression is `r summary(lm(observations ~ population, data=num_vs_pop))$r.squared`.

@@

gplot(num_vs_pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Observations (log)") +
    ggtitle("Observations per capita")
```

The log number of observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.

@@

gplot(num_vs_pop_rasterized_LM) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Observations (log)") +
    ggtitle("Observations corrected for inhabitants")
```

The log number of observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway, divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation), rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.

## Bioclimatic variables

```{r climate_data, messages=FALSE, echo=FALSE}

env <- getData('worldclim', var='bio', res=5)

env2 <- projectRaster(env[[1]], grid, method="bilinear")

# plot(env, 1, main=NULL, axes=FALSE)

gplot(env2) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
    geom_tile(aes(fill=value/10), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA, direction=-1) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "°C") +
    ggtitle("Annual Mean Temperature")

The Annual Mean Temperature in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.
